\thetitle

### 2020-11-02 sun_phases/spin_qudit_tomo.tex
major changes to the section on the reconstruction error bound, which now constructs a 'measurement matrix' and computes its singular values (rather than constructing an 'overlap matrix' and computing its eigenvalues)

\newcommand{\B}{\mathcal{B}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\Y}{\mathcal{Y}}

@@

A set of measurement axes $V=\set{\v v}$ nominally induces a set of projectors $\set{\Pi_{\v v\mu}^V}$ that will be measured in an experiment, where for clarify we index objects by $V$ to emphasize their dependence on the choice of measurement axes.
By a simple change of basis, measuring these projectors is equivalent to measuring the set of transition operators $\v T_V\equiv\set{T_{\v v\ell,0}}$.
The operators in $\v T_V$ can be used to construct the overlap (Gram) matrix
\begin{align}
  G_V \equiv \sum_{\v v,\v v',\ell}
  \obk{T_{\v v\ell,0}|T_{\v v'\ell,0}} \op{\v v\ell}{\v v'\ell},
  \label{eq:overlap_matrix}
\end{align}
where we neglect cross terms with different degrees $\ell$ because the corresponding inner products vanish, i.e.~$\obk{T_{\v v\ell,0}|T_{\v v'\ell',0}}=0$ for all $\ell\ne\ell'$.
If the operators in $\v T_V$ span the entire ($d^2$-dimensional) space of operators on a $d$-level spin qudit, which is a necessary and sufficient condition for $V$ to allow for full state tomography, then $G_V$ must have exactly $d^2$ nonzero eigenvalues.
Indexing these eigenvalues $G^V_k$ and the corresponding (normalized) eigenvectors $\v x^V_k\equiv(x^V_{k,1},x^V_{k,2},\cdots)$ by an integer $k\in\set{1,2,\cdots,d^2}$, we can construct the orthonormal operators
\begin{align}
  Q^V_k \equiv \sum_j g^V_{kj} T_j,
  &&
  g^V_{kj} \equiv \f{x^V_{kj}}{\sqrt{G^V_k}},
\end{align}
where for shorthand we use a combined index $j=\p{\v v_j,\ell_j}$ to specify both a measurement axis $\v v_j$ and a degree $\ell_j$, which identify the transition operator $T_j\equiv T_{\v v_j\ell_j,0}$.
These operators allow us to expand any state $\rho$ of a $d$-level spin qudit in the form
\begin{align}
  \rho = \sum_{k=1}^{d^2} \rho_k^V Q_k^V,
  &&
  \rho_k^V \equiv \bk{Q^V_k}_\rho.
\end{align}
Given empirical estimates $E_j$ of the expectation values $\bk{T_j}_\rho$, an empirical estimate $\tilde\rho_V$ of $\rho$ is then
\begin{align}
  \tilde\rho_V \equiv \sum_k \tilde\rho^V_k Q^V_k,
  \label{eq:reconstructed_state}
\end{align}
where
\begin{align}
  \tilde\rho^V_k \equiv \sum_j g^V_{kj} E_j
  \approx \sum_j g^V_{kj} \bk{T_j}_\rho
  = \bk{Q^V_k}_\rho
  = \rho^V_k.
\end{align}
The eigenvalues of the overlap matrix $G_V$ allow us to make concrete statements about the statistical error between the empirical estimate $\tilde\rho_V$ and the true state $\rho$.
Assume, for example, that the estimates $E_j$ are equal to $\bk{T_j}_\rho$ up to Gaussian noise with variance at most $\epsilon^2$, i.e.
\begin{align}
  E_j = \bk{T_j}_\rho + \epsilon_j,
  &&
  \bbk{\epsilon_j \epsilon_{j'}} \le \epsilon^2 \delta_{jj'},
\end{align}
where $\set{\epsilon_j}$ are independent random variables, and we use the double brackets $\bbk{\cdot}$ to denote statistical averaging.
The mean squared error with which $\tilde\rho^V_k$ approximates $\rho^V_k$ is then
\begin{align}
  \Bbk{\abs{\tilde\rho^V_k-\rho^V_k}^2}
  &= \Bbk{\p{\tilde\rho^V_k-\rho^V_k}^*\p{\tilde\rho^V_k-\rho^V_k}} \\
  &= \sum_{j,j'} \p{g^V_{kj}}^* g^V_{kj'}\,
  \bbk{\epsilon_j \epsilon_{j'}} \\
  &\le \sum_j \abs{g^V_{kj}}^2 \epsilon^2
  = \f{\epsilon^2}{G^V_k}.
\end{align}
Using the fact that the operators $Q^V_k$ are orthonormal, we can therefore bound the mean squared 2-norm (Euclidean) distance between $\tilde\rho_V$ and $\rho$ as
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}_2^2}
  = \sum_k \Bbk{\abs{\tilde\rho^V_k - \rho^V_k}^2}
  \le \epsilon^2 \M_V^2,
  \label{eq:bound}
\end{align}
where
\begin{align}
  \M_V^2 \equiv \sum_k \f1{G^V_k}
  \label{eq:scale}
\end{align}
is a scale determined by the set of measurement axes $V$.
Note that the sum in Eq.~\eqref{eq:scale} is performed over the $d^2$ largest eigenvalues of $G_V$, and therefore may be divergent if $G_V$ has fewer than $d^2$ nonzero eigenvalues.

The complexity of computing $\M_V^2$ is determined by the need to:
\begin{enumerate*}
\item compute all rotated transition operators $T_{\v v\ell,0}$, which by the expansion $T_{\v v\ell,0}=R\p{\v v}T_{\ell,0}R\p{\v v}^\dag$ takes $O(\abs{V}d^4)$ serial or $O(d^3)$ parallel runtime;
\item construct the overlap matrix $G_V$, whose block diagonal structure makes possible in $O(\abs{V}^2d^3)$ serial or $O(\abs{V}^2d^2)$ parallel runtime; and
\item diagonalize the blocks of $G_V$, which takes $O(\abs{V}^3d)$ serial or $O(\abs{V}^3)$ parallel runtime.
\end{enumerate*}
Altogether, assuming $\abs{V}\propto d$ the bound in Eq.~\eqref{eq:bound} can be certified in $O(d^5)$ serial or $O(d^4)$ parallel runtime.
In practice, for modest values of $d$ we find that vectorized operations can reduce serial runtime to $\sim O(r^4)$, and that serial execution time is divided about evenly between computing rotated transition operators $T_{\v v\ell,0}$ and constructing the overlap matrix $G_V$, with a negligible amount of time spent on diagonalization (see Figure [\red{REF}]).  [\red{todo: add figure showing runtime for each routine as a function of $d$ with $\abs{V}=2d-1$; modify text appropriately after making the figure}].
We suspect that more efficient representations of the transition operators $T_{\ell m}$, namely as $\p{2\ell+1}$-component vectors rather than $d\times d$ matrices, together with appropriate changes to the representation of the adjoint rotation operator $\R\p{\v v}$, should reduce the computational complexity of computing $\M_V^2$ by a factor of $O(d)$.

@@

The ability to certify a statistical error bound on the empirical estimate $\tilde\rho_V$ of an unknown quantum state $\rho$ motivates the following simple protocol for spin qudit tomography: select a random set of measurement axes $V$ by uniformly sampling points on the sphere \footnote{To sample a point $\p{\alpha,\beta}$ from the uniform distribution on the sphere (with azimuthal angle $\alpha$ and polar angle $\beta$), you can sample a point $\p{a,b}\in[0,1]\times[0,1]$ from the uniform distribution on the unit square, and then set $\alpha=2\pi a$ and $\beta=\arccos\p{1-2b}$.}, and variationally optimize the $2\abs{V}$ parameters in $V$ (two angles for each point $\v v\in V$) to minimize the error scale $\M_V$ defined in Eq.~\eqref{eq:scale}.
If $\abs{V}$ is too large for variational optimization to be practical, one can instead generate many random sets of measurement axes, and then choose the set that minimizes $\M_V$.
After choosing a set of measurement axes $V$, experimentally characterize the expectation values $\bk{T_j}_\rho$.
If the experimental measurements $E_j$ of $\bk{T_j}_\rho$ are accurate to within an error $\lesssim\epsilon$, then the state $\tilde\rho_V$ defined in Eq.~\eqref{eq:reconstructed_state} is an approximation of the true state $\rho$ to within a 2-norm distance $\lesssim\epsilon\M_V$.
Furthermore, if the empirical estimate $\tilde\rho_V$ has negative eigenvalues, its 2-norm distance from $\rho$ can be further reduced with maximum-likelihood corrections \cite{smolin2012efficient}.

@@

As a last point, we mention that the error scale $\M_V$ can also evaluate the ``quality'' of measurement axes $V$ chosen with existing methods for sparse signal recovery on the sphere \cite{rauhut2011sparse, alem2012sparse}.
In practice, we find that uniform random sampling outperforms these methods, in the sense that it generates low-error sets of measurement axes with higher probability.
Two observations may explain the poor performance of these sparse signal recovery methods.
First, spin qudit tomography essentially boils down to recovering sparse signals with a particular harmonic structure, namely a fixed degree $\ell$.
The existing sparse signal recovery algorithms do not assume or account for any harmonic structure, and may therefore perform worse than expected on this particular subset of signals.
Second, spin qudit tomography requires recovering signals that are band-limited at degree $d$, but have $O(d)$ non-zero harmonic coefficients (out of $d^2$ coefficients in total).
The existing sparse signal recovery methods, on the other hand, may only perform well for signals with sparsity much smaller than $d$.
In any case, the fact that uniform sampling outperforms these methods only simplifies our randomized tomography protocol.

@@

We conjecture, and numerically verify for $d\le200$, that measurements of spin projection along $2d-1$ axes suffice to reconstruct an arbitrary state of a $d$-level spin qudit (see Figure [\red{REF}]).
Even so, $2d-1$ measurement axes may not be the ``optimal'' number to use.
Increasing the number of measurement axes generally decreases the error scale $\M_V$, but at the cost of having to estimate more observables.
If we fix the total number of measurements (or experimental trials), then having to estimate more observables means that fewer measurements are devoted to estimating any particular observable, which in turn increases the error $\epsilon$ on those estimates.
This trade-off begs the question: how many measurement axes should be chosen to perform spin qudit tomography?

The reconstruction error bound in Eq.~\eqref{eq:bound} provides a straightforward answer: the number of measurement axes should be chosen to minimize the product $\epsilon^2\M_V^2$, which bounds the expected error in the estimate $\tilde\rho_V$ of an unknown state $\rho$.
If there are a total of $N$ spin projection measurements onto $r$ quantization axes, then $\sim N/r$ measurements are used to estimate spin projection onto any given axis.
The variance in estimates of spin projection is then $\epsilon^2\sim r/N$, or simply $\epsilon^2\sim r$ at a fixed total number of measurements $N$.
Altogether, the number of measurement axes should be chosen in such a way as to minimize $r\M_V^2$ with $\abs{V}=r$.

>>>>

\newcommand{\B}{\mathcal{B}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\R}{\mathcal{R}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\Y}{\mathcal{Y}}

@@

A set of measurement axes $V=\set{\v v}$ nominally induces a set of projectors $\set{\Pi_{\v v\mu}}$ that will be measured in an experiment.
By a simple change of basis, measuring these projectors is equivalent to measuring the set of transition operators $\v T_V\equiv\set{T_{\v v\ell,0}}$.
By flattening qudit operators $X = \sum_{\mu,\nu} X_{\mu\nu} \op{\mu}{\nu}$ into the $d^2$-component vectors $\oket{X} \equiv \sum_{\mu,\nu} X_{\mu\nu} \ket{\mu\nu}$, we can construct the {\it measurement matrix}
\begin{align}
  M_V \equiv \sum_{\v v,\ell} \oket{T_{\v v\ell,0}} \bra{\v v\ell},
  \label{eq:meas_mat}
\end{align}
whose columns are flattenings of the operators in $\v T_V$.
A necessary and sufficient for $V$ to allow for full state tomography is for the operators in $\v T_V$, or equivalently the columns of $M_V$, to span the entire ($d^2$-dimensional) space of operators on a $d$-level spin qudit.
In this case, $M_V$ must be full rank, with $d^2$ nonzero singular values.
Indexing these singular values $M^V_k$ and the corresponding (normalized) right singular vectors $\v x^V_k\equiv(x^V_{k,1},x^V_{k,2},\cdots)$ by an integer $k\in\set{1,2,\cdots,d^2}$, we can construct the orthonormal qudit operators
\begin{align}
  Q^V_k \equiv \sum_j q^V_{kj} T_j,
  &&
  q^V_{kj} \equiv \f{x^V_{kj}}{M^V_k},
\end{align}
where for shorthand we use a combined index $j=\p{\v v_j,\ell_j}$ to specify both a measurement axis $\v v_j$ and a degree $\ell_j$, which identify the transition operator $T_j\equiv T_{\v v_j\ell_j,0}$.
These operators allow us to expand any state $\rho$ of a $d$-level spin qudit in the form
\begin{align}
  \rho = \sum_{k=1}^{d^2} \rho_k^V Q_k^V,
  &&
  \rho_k^V \equiv \bk{Q^V_k}_\rho.
\end{align}
Given empirical estimates $E_j$ of the expectation values $\bk{T_j}_\rho$, an empirical estimate $\tilde\rho_V$ of $\rho$ is then
\begin{align}
  \tilde\rho_V \equiv \sum_k \tilde\rho^V_k Q^V_k,
  \label{eq:reconstructed_state}
\end{align}
where
\begin{align}
  \tilde\rho^V_k \equiv \sum_j q^V_{kj} E_j
  \approx \sum_j q^V_{kj} \bk{T_j}_\rho
  = \bk{Q^V_k}_\rho
  = \rho^V_k.
\end{align}
The singular values of $M_V$ allow us to make concrete statements about the statistical error between the empirical estimate $\tilde\rho_V$ and the true state $\rho$.
Assume, for example, that the estimates $E_j$ are equal to $\bk{T_j}_\rho$ up to Gaussian noise with variance at most $\epsilon^2$, i.e.
\begin{align}
  E_j = \bk{T_j}_\rho + \epsilon_j,
  &&
  \bbk{\epsilon_j \epsilon_{j'}} \le \epsilon^2 \delta_{jj'},
\end{align}
where $\set{\epsilon_j}$ are independent random variables, and we use the double brackets $\bbk{\cdot}$ to denote statistical averaging.
The mean squared error with which $\tilde\rho^V_k$ approximates $\rho^V_k$ is then
\begin{align}
  \Bbk{\abs{\tilde\rho^V_k-\rho^V_k}^2}
  &= \Bbk{\p{\tilde\rho^V_k-\rho^V_k}^*\p{\tilde\rho^V_k-\rho^V_k}} \\
  &= \sum_{j,j'} \p{q^V_{kj}}^* q^V_{kj'}\,
  \bbk{\epsilon_j \epsilon_{j'}} \\
  &\le \sum_j \abs{q^V_{kj}}^2 \epsilon^2
  = \p{\f{\epsilon}{M^V_k}}^2.
\end{align}
Using the fact that the operators $Q^V_k$ are orthonormal, we can therefore bound the mean squared 2-norm (Euclidean) distance between $\tilde\rho_V$ and $\rho$ as
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}_2^2}
  = \sum_k \Bbk{\abs{\tilde\rho^V_k - \rho^V_k}^2}
  \le \epsilon^2 \S_V^2,
  \label{eq:bound}
\end{align}
where
\begin{align}
  \S_V \equiv \norm{M_V^{-1}}_2 = \sqrt{\sum_k\f1{\p{M^V_k}^2}},
  \label{eq:scale}
\end{align}
which we call the {\it error scale} of $V$, is the 2-norm of the pseudoinverse $M_V^{-1}$ of $M_V$.
Note that the bound in Eq.~\eqref{eq:bound} only holds if the measurement matrix $M_V$ is nonsingular, as otherwise some of the operators $Q^V_k$ that we used to expand $\rho$ and $\tilde\rho_V$ are undefined, reflecting the fact that if $M_V$ is singular then measuring spin projections along all axes in $V$ does not provide sufficient information to reconstruct arbitrary quantum states.

The complexity of computing the error scale $\S_V$ is nominally determined by the need to
\begin{enumerate*}
\item construct the measurement matrix $M_V$, and
\item compute the singular values of $M_V$.
\end{enumerate*}
By the expansion $T_{\v v\ell,0} = R\p{\v v} T_{\ell,0} R\p{\v v}^\dag$, each of the $\abs{V}d$ columns of $M_V$ takes $O(d^3)$ time to compute, for an overall $O(\abs{V}d^4)$ serial or $O(d^3)$ parallel runtime.
In order to compute the singular values of $M_V$, we first note that the orthogonality of transition operators with different degrees, i.e.~$\obk{T_{\v v\ell,0}|T_{\v v',\ell',0}} = 0$ for $\ell\ne\ell'$, implies that $M_V$ can be decomposed into a direct sum of the rectangular blocks
\begin{align}
  M_{V\ell} \equiv \sum_{\v v} \oket{T_{\v v\ell,0}} \bra{\v v},
\end{align}
using which we can expand
\begin{align}
  \S_V^2 = \sum_\ell \S_{V\ell}^2,
  &&
  \S_{V\ell} \equiv \norm{M_{V\ell}^{-1}}_2,
\end{align}
where $M_{V\ell}^{-1}$ is the pseudoinverse of $M_{V\ell}$.
Finding the singular values of each block $M_{V\ell}$ takes $O(\abs{V}^2d^2)$ time at most \footnote{The calculation of $\S_{V\ell}$ can be sped up by the fact that it is independent of $d$ as long as $d\ge\ell+1$, so we can essentially set $d=\ell+1$ for the purposes of computing $\S_{V\ell}$}.
Altogether, if we assume that $\abs{V}\sim d$ then certifying the bound in Eq.~\eqref{eq:bound} by computing $\S_V$ takes $O(d^5)$ serial or $O(d^4)$ parallel runtime (see Figure [\red{REF}]).  [\red{todo: add figure showing runtime for each routine as a function of $d$ with $\abs{V}=2d-1$; modify text appropriately after making the figure}].
We suspect that more efficient representations of the transition operators $T_{\ell m}$, namely as $\p{2\ell+1}$-component vectors rather than $d\times d$ matrices, together with appropriate changes to the representation of the adjoint rotation operator $\R\p{\v v}$, can reduce the complexity of computing $\S_V$ by a factor of $O(d)$.

@@

The ability to certify a statistical error bound on the empirical estimate $\tilde\rho_V$ of an unknown quantum state $\rho$ motivates the following simple protocol for spin qudit tomography: select a random set of measurement axes $V$ by uniformly sampling points on the sphere \footnote{To sample a point $\p{\alpha,\beta}$ from the uniform distribution on the sphere (with azimuthal angle $\alpha$ and polar angle $\beta$), you can sample a point $\p{a,b}\in[0,1]\times[0,1]$ from the uniform distribution on the unit square, and then set $\alpha=2\pi a$ and $\beta=\arccos\p{1-2b}$.}, and variationally optimize the $2\abs{V}$ parameters in $V$ (two angles for each point $\v v\in V$) to minimize the error scale $\S_V$ defined in Eq.~\eqref{eq:scale}.
If $\abs{V}$ is too large for variational optimization to be practical, one can instead generate many random sets of measurement axes, and then choose the set that minimizes $\S_V$.
After choosing a set of measurement axes $V$, experimentally characterize the expectation values $\bk{T_j}_\rho$.
If the experimental measurements $E_j$ of $\bk{T_j}_\rho$ are accurate to within an error $\lesssim\epsilon$, then the state $\tilde\rho_V$ defined in Eq.~\eqref{eq:reconstructed_state} is an approximation of the true state $\rho$ to within a 2-norm distance $\lesssim\epsilon\S_V$.
Furthermore, if the empirical estimate $\tilde\rho_V$ has negative eigenvalues, its 2-norm distance from $\rho$ can be further reduced with maximum-likelihood corrections \cite{smolin2012efficient}.

@@

As a last point, we mention that the error scale $\S_V$ can also evaluate the ``quality'' of measurement axes $V$ chosen with existing methods for sparse signal recovery on the sphere \cite{rauhut2011sparse, alem2012sparse}.
In practice, we find that uniform random sampling outperforms these methods, in the sense that it generates low-error-scale sets of measurement axes with higher probability.
Two observations may explain the poor performance of these sparse signal recovery methods.
First, spin qudit tomography essentially boils down to recovering sparse signals with a particular harmonic structure, namely a fixed degree $\ell$.
The existing sparse signal recovery algorithms do not assume or account for any harmonic structure, and may therefore perform worse than expected on this particular subset of signals.
Second, spin qudit tomography requires recovering signals that are band-limited at degree $d$, but have $O(d)$ non-zero harmonic coefficients (out of $d^2$ coefficients in total).
The existing sparse signal recovery methods, on the other hand, may only perform well for signals with sparsity much smaller than $d$.
In any case, the fact that uniform sampling outperforms these methods only simplifies our randomized tomography protocol.

@@

We conjecture, and numerically verify for $d\le200$, that measurements of spin projection along $2d-1$ axes suffice to reconstruct an arbitrary state of a $d$-level spin qudit (see Figure [\red{REF}]).
Even so, $2d-1$ measurement axes may not be the ``optimal'' number to use.
Increasing the number of measurement axes generally decreases the error scale $\S_V$, but at the cost of having to estimate more observables.
If we fix the total number of measurements (or experimental trials), then having to estimate more observables means that fewer measurements are devoted to estimating any particular observable, which in turn increases the error $\epsilon$ on those estimates.
This trade-off begs the question: how many measurement axes should be chosen to perform spin qudit tomography?

The reconstruction error bound in Eq.~\eqref{eq:bound} provides a straightforward answer: the number of measurement axes should be chosen to minimize the product $\epsilon^2\S_V^2$, which bounds the expected error in the estimate $\tilde\rho_V$ of an unknown state $\rho$.
If there are a total of $N$ spin projection measurements onto $r$ quantization axes, then $\sim N/r$ measurements are used to estimate spin projection onto any given axis.
The variance in estimates of spin projection is then $\epsilon^2\sim r/N$, or simply $\epsilon^2\sim r$ at a fixed total number of measurements $N$.
Altogether, the number of measurement axes should be chosen in such a way as to minimize $r\S_V^2$ with $\abs{V}=r$.
